---
title: "2026-09-10_UMR_peak_analysis"
author: "Guilherme V Yoshicawa"
date: "`r Sys.Date()`"
output: html_document
---

# Notes
-

# Setup
```{r setup, include=FALSE}
knitr::opts_knit$set(
  root.dir = "R:/YOSHI2026_EPIALLELES-Q9486/data/Sb-UMRseq-diversity-panel/analysis"
)
# run at the start of every R session to set up the environment
source(Sys.getenv("VSCODE_INIT_R"), chdir = TRUE)
.First.sys()
Sys.setenv(RSTUDIO_PANDOC="C:/Program Files/RStudio/resources/app/bin/quarto/bin/tools")
# install.packages(c("tidyverse", "viridis"))
library("rmarkdown")
library("viridis")
library(tidyverse)
library(pandoc)
```

``` {r}
## Load files (working directory comes from the setup chunk above)
bp50 <-  read.csv("epic2_50_region_metrics.tsv", sep = "\t", header = TRUE)
head(bp50)
bp100 <-  read.csv("epic2_100_region_metrics.tsv", sep = "\t", header = TRUE)
head(bp100)
# NB: these two are comma-separated (not tab), with the counts quoted and
# thousands-separated, e.g. "13,804,781" - so they need sep = "," and the
# commas stripped before as.numeric(), or every count comes back NA.
read_counts <-  read.csv("UMR_libraries_reads.csv", sep = ",", header = TRUE)
low_read    <-  read.csv("low_depth_samples.csv",   sep = ",", header = TRUE)

#samples with reads under 9M will be excluded (9 samples)

# both files carry "Sample" / "Paired reads"; normalise to sample / paired_reads
parse_reads <- function(df) {
  df %>%
    rename(sample = 1, paired_reads = 2) %>%
    mutate(paired_reads = as.numeric(gsub(",", "", as.character(paired_reads))))
}
read_counts <- parse_reads(read_counts)
low_read    <- parse_reads(low_read)

head(read_counts)
nrow(low_read)   # samples excluded for low depth
```

# Combine both bin sizes into one long table
```{r}
metrics <- bind_rows(
  bp50  %>% mutate(bin_size = "50 bp"),
  bp100 %>% mutate(bin_size = "100 bp")
) %>%
  mutate(
    bin_size = factor(bin_size, levels = c("50 bp", "100 bp")),
    # guard: samples with empty fields make read.csv return character columns
    across(c(MEDIAN_LENGTH, MEAN_LENGTH, NUM_REGIONS), as.numeric)
  ) %>%
  pivot_longer(
    cols      = c(MEDIAN_LENGTH, MEAN_LENGTH, NUM_REGIONS),
    names_to  = "metric",
    values_to = "value"
  ) %>%
  filter(!is.na(value)  # drops samples with empty/NA metrics
  ) %>%
  filter(!sample %in% low_read$sample) %>%  # drops samples with low read depth
  # append the library size so metrics can be modelled against sequencing depth
  left_join(read_counts, by = "sample") %>%
  mutate(reads_M = paired_reads / 1e6)

# sanity check: every retained sample should have a read count
stopifnot(!any(is.na(metrics$paired_reads)))

# per-metric, per-bin summary statistics used for the annotations
metric_stats <- metrics %>%
  group_by(metric, bin_size) %>%
  summarise(
    n      = n(),
    mean   = mean(value),
    median = median(value),
    sd     = sd(value),
    .groups = "drop"
  )
metric_stats
```

# Summary statistics: 50 bp vs 100 bp
```{r}
# full descriptive table (mean, sd, se, CV, quartiles, range) per metric x bin size
summary_stats <- metrics %>%
  group_by(metric, bin_size) %>%
  summarise(
    n      = n(),
    mean   = mean(value),
    sd     = sd(value),
    se     = sd(value) / sqrt(n()),
    cv_pct = 100 * sd(value) / mean(value),
    min    = min(value),
    q1     = quantile(value, 0.25),
    median = median(value),
    q3     = quantile(value, 0.75),
    max    = max(value),
    .groups = "drop"
  ) %>%
  mutate(across(where(is.numeric) & !c(n), ~ round(.x, 2))) %>%
  arrange(metric, bin_size)

summary_stats

# compact "mean ± sd" view, one row per metric
summary_stats %>%
  mutate(`mean ± sd` = paste0(format(mean, big.mark = ",", trim = TRUE),
                              " ± ", format(sd, big.mark = ",", trim = TRUE))) %>%
  select(metric, bin_size, n, `mean ± sd`, median) %>%
  pivot_wider(names_from = bin_size, values_from = c(n, `mean ± sd`, median))
```

# Density histogram function
```{r}
plot_metric <- function(which_metric, x_lab, digits = 0,
                        dat = metrics, stat_tbl = metric_stats) {

  d <- filter(dat, metric == which_metric)
  s <- filter(stat_tbl, metric == which_metric)

  # multi-line stats label, one line per bin size
  stat_label <- paste0(
    s$bin_size, ": mean = ", format(round(s$mean, digits), big.mark = ",", trim = TRUE),
    ", median = ", format(round(s$median, digits), big.mark = ",", trim = TRUE),
    " (n = ", s$n, ")",
    collapse = "\n"
  )

  ggplot(d, aes(x = value, fill = bin_size, colour = bin_size)) +
    geom_histogram(
      aes(y = after_stat(density)),
      bins = 30, position = "identity",
      alpha = 0.4, linewidth = 0
    ) +
    geom_density(fill = NA, linewidth = 0.3) +
    geom_vline(
      data = s, aes(xintercept = mean, colour = bin_size),
      linetype = "dashed", linewidth = 0.25, inherit.aes = FALSE
    ) +
    scale_x_continuous(labels = scales::comma) +
    scale_fill_viridis_d(option = "plasma", direction = -1, begin = 0.15, end = 0.75) +
    scale_colour_viridis_d(option = "plasma", direction = -1, begin = 0.15, end = 0.75) +
    labs(
      title  = NULL,
      x      = x_lab,
      y      = "Density",
      fill   = "Bin size",
      colour = "Bin size"
    ) +
    # extra headroom so the annotation never collides with the density curves
    scale_y_continuous(expand = expansion(mult = c(0.02, 0.22))) +
    annotate(
      "text", x = Inf, y = Inf, hjust = 1.05, vjust = 1.3,
      label = stat_label,
      size = 2.5, lineheight = 1.15
    ) +
    theme_classic() +
    theme(
      axis.text.x     = element_text(size = 7, colour = "black"),
      axis.text.y     = element_text(size = 7, colour = "black"),
      axis.title      = element_text(size = 8, colour = "black"),
      axis.line       = element_line(linewidth = 0.25, colour = "black"),
      axis.ticks      = element_line(linewidth = 0.25, colour = "black"),
      legend.text     = element_text(size = 7),
      legend.title    = element_text(size = 8, face = "bold"),
      legend.key.size = unit(0.35, "cm")
    )
}
```

# Median region length
```{r}
a <- plot_metric("MEDIAN_LENGTH", "Median region length (bp)")
a
```

# Mean region length
```{r}
#outlier on 3kb only has frip of 84, and 17M reads - check IGV wig files
#Sorghum_250_FF_SC557-14E
b <- plot_metric("MEAN_LENGTH", "Mean region length (bp)")
b
```

# Number of regions
```{r}
c <- plot_metric("NUM_REGIONS", "Number of regions")
c
```

```{r, include=FALSE}
ggsave("UMR_density_median_length.png", a, width = 14, height = 9, units = "cm")
ggsave("UMR_density_mean_length.png",   b, width = 14, height = 9, units = "cm")
ggsave("UMR_density_num_regions.png",   c, width = 14, height = 9, units = "cm")
```

```{r}
frip50 <-  read.csv("FRIP_50_scraped.tsv", sep = "\t", header = TRUE)
head(frip50)
frip100 <-  read.csv("FRIP_100_scraped.tsv", sep = "\t", header = TRUE)
head(frip100)
```

# Combine FRIP scores from both bin sizes
```{r}
# header in these files is "sample<TAB>FRIP score", so read.csv names it
# FRIP.score - select by position to stay robust to that
frip <- bind_rows(
  frip50  %>% mutate(bin_size = "50 bp"),
  frip100 %>% mutate(bin_size = "100 bp")
) %>%
  rename(FRIP = 2) %>%
  mutate(
    bin_size = factor(bin_size, levels = c("50 bp", "100 bp")),
    FRIP     = as.numeric(FRIP)
  ) %>%
  pivot_longer(cols = FRIP, names_to = "metric", values_to = "value") %>%
  filter(!is.na(value)) %>%                 # drops samples with empty/NA FRIP
  filter(!sample %in% low_read$sample) %>%  # same low-depth exclusion as metrics
  # append the library size so FRIP can also be checked against depth
  left_join(read_counts, by = "sample") %>%
  mutate(reads_M = paired_reads / 1e6)

stopifnot(!any(is.na(frip$paired_reads)))

frip_stats <- frip %>%
  group_by(metric, bin_size) %>%
  summarise(
    n      = n(),
    mean   = mean(value),
    median = median(value),
    sd     = sd(value),
    .groups = "drop"
  )
frip_stats
```

# FRIP summary statistics: 50 bp vs 100 bp
```{r}
frip_summary <- frip %>%
  group_by(bin_size) %>%
  summarise(
    n      = n(),
    mean   = mean(value),
    sd     = sd(value),
    se     = sd(value) / sqrt(n()),
    cv_pct = 100 * sd(value) / mean(value),
    min    = min(value),
    q1     = quantile(value, 0.25),
    median = median(value),
    q3     = quantile(value, 0.75),
    max    = max(value),
    .groups = "drop"
  ) %>%
  mutate(across(where(is.numeric) & !c(n), ~ round(.x, 2)))

frip_summary

# compact "mean ± sd" view
frip_summary %>%
  mutate(`mean ± sd` = paste0(mean, " ± ", sd)) %>%
  select(bin_size, n, `mean ± sd`, median, min, max)
```

# Paired comparison of FRIP between bin sizes
```{r}
frip_paired <- frip %>%
  pivot_wider(names_from = bin_size, values_from = value,
              names_prefix = "bin_") %>%
  filter(!is.na(`bin_50 bp`), !is.na(`bin_100 bp`)) %>%
  mutate(diff = `bin_50 bp` - `bin_100 bp`)

frip_tests <- frip_paired %>%
  summarise(
    n_pairs    = n(),
    mean_50    = mean(`bin_50 bp`),
    sd_50      = sd(`bin_50 bp`),
    mean_100   = mean(`bin_100 bp`),
    sd_100     = sd(`bin_100 bp`),
    mean_diff  = mean(diff),
    sd_diff    = sd(diff),
    pct_change = 100 * mean(diff) / mean(`bin_100 bp`),  # 50 bp vs 100 bp
    t_p        = t.test(`bin_50 bp`, `bin_100 bp`, paired = TRUE)$p.value,
    wilcox_p   = wilcox.test(`bin_50 bp`, `bin_100 bp`, paired = TRUE)$p.value,
    cohens_dz  = mean(diff) / sd(diff),                  # paired effect size
    r_pearson  = cor(`bin_50 bp`, `bin_100 bp`)          # agreement between bins
  ) %>%
  mutate(across(where(is.numeric) & !c(n_pairs), ~ signif(.x, 3)))

frip_tests
```

# FRIP density histogram
```{r}
d <- plot_metric("FRIP", "FRIP score (%)", digits = 2,
                 dat = frip, stat_tbl = frip_stats)
d
```

# Save FRIP figure
```{r}
ggsave("UMR_density_FRIP.png", d, width = 14, height = 9, units = "cm")
```

# Region metrics vs sequencing depth
```{r}
# how much of the variation in each metric is explained by library size?
depth_cor <- metrics %>%
  group_by(metric, bin_size) %>%
  summarise(
    n            = n(),
    slope        = coef(lm(value ~ reads_M))[2],   # change per million reads
    r_pearson    = cor(reads_M, value),
    r2           = cor(reads_M, value)^2,
    rho_spearman = cor(reads_M, value, method = "spearman"),
    p_pearson    = cor.test(reads_M, value)$p.value,
    .groups = "drop"
  ) %>%
  mutate(
    p_adj = p.adjust(p_pearson, method = "BH"),
    across(where(is.numeric) & !c(n), ~ signif(.x, 3))
  ) %>%
  arrange(metric, bin_size)

depth_cor
```

# Scatter function: metric vs read count
```{r}
plot_vs_depth <- function(which_metric, y_lab,
                          dat = metrics, stat_tbl = depth_cor) {

  d <- filter(dat, metric == which_metric)
  s <- filter(stat_tbl, metric == which_metric)

  stat_label <- paste0(
    s$bin_size, ": R² = ", format(round(s$r2, 2), nsmall = 2),
    ", rho = ", format(round(s$rho_spearman, 2), nsmall = 2),
    collapse = "\n"
  )

  ggplot(d, aes(x = reads_M, y = value, colour = bin_size, fill = bin_size)) +
    geom_point(size = 0.5, alpha = 0.6, stroke = 0) +
    geom_smooth(method = "lm", se = TRUE, linewidth = 0.4, alpha = 0.15) +
    scale_x_continuous(labels = scales::comma) +
    scale_y_continuous(labels = scales::comma,
                       expand = expansion(mult = c(0.02, 0.18))) +
    scale_fill_viridis_d(option = "plasma", direction = -1,
                         begin = 0.15, end = 0.75) +
    scale_colour_viridis_d(option = "plasma", direction = -1,
                           begin = 0.15, end = 0.75) +
    labs(
      title  = NULL,
      x      = "Paired reads (millions)",
      y      = y_lab,
      fill   = "Bin size",
      colour = "Bin size"
    ) +
    annotate(
      "text", x = Inf, y = Inf, hjust = 1.05, vjust = 1.3,
      label = stat_label,
      size = 2.5, lineheight = 1.15
    ) +
    theme_classic() +
    theme(
      axis.text.x     = element_text(size = 7, colour = "black"),
      axis.text.y     = element_text(size = 7, colour = "black"),
      axis.title      = element_text(size = 8, colour = "black"),
      axis.line       = element_line(linewidth = 0.25, colour = "black"),
      axis.ticks      = element_line(linewidth = 0.25, colour = "black"),
      legend.text     = element_text(size = 7),
      legend.title    = element_text(size = 8, face = "bold"),
      legend.key.size = unit(0.35, "cm")
    )
}
```

# Median region length vs read count
```{r}
e <- plot_vs_depth("MEDIAN_LENGTH", "Median region length (bp)")
e
```

# Mean region length vs read count
```{r}
f <- plot_vs_depth("MEAN_LENGTH", "Mean region length (bp)")
f
```

# Number of regions vs read count
```{r}
g <- plot_vs_depth("NUM_REGIONS", "Number of regions")
g
```

# Save depth figures
```{r, include=FALSE}
ggsave("UMR_depth_median_length.png", e, width = 14, height = 9, units = "cm")
ggsave("UMR_depth_mean_length.png",   f, width = 14, height = 9, units = "cm")
ggsave("UMR_depth_num_regions.png",   g, width = 14, height = 9, units = "cm")
```

# FRIP vs sequencing depth
```{r}
# same correlation table as depth_cor, built on the FRIP data
frip_depth_cor <- frip %>%
  group_by(metric, bin_size) %>%
  summarise(
    n            = n(),
    slope        = coef(lm(value ~ reads_M))[2],   # FRIP % per million reads
    r_pearson    = cor(reads_M, value),
    r2           = cor(reads_M, value)^2,
    rho_spearman = cor(reads_M, value, method = "spearman"),
    p_pearson    = cor.test(reads_M, value)$p.value,
    .groups = "drop"
  ) %>%
  mutate(
    p_adj = p.adjust(p_pearson, method = "BH"),
    across(where(is.numeric) & !c(n), ~ signif(.x, 3))
  )

frip_depth_cor

h <- plot_vs_depth("FRIP", "FRIP score (%)",
                   dat = frip, stat_tbl = frip_depth_cor)
h
```

```{r,include=FALSE}
ggsave("UMR_depth_FRIP.png", h, width = 14, height = 9, units = "cm")
```

#find frip score of removed samples
```{r}
frip_removed <- frip50 %>%
  rename(FRIP = 2) %>%                                  # header is "FRIP.score"
  mutate(FRIP = as.numeric(FRIP)) %>%
  left_join(read_counts, by = "sample") %>%             # depth for every sample
  mutate(
    reads_M   = paired_reads / 1e6,
    low_depth = sample %in% low_read$sample             # TRUE = excluded sample
  ) %>%
  arrange(FRIP)

# the excluded samples on their own
frip_removed %>% filter(low_depth)

# do low-depth samples have lower FRIP?
frip_removed %>%
  group_by(low_depth) %>%
  summarise(n = n(), median_FRIP = median(FRIP, na.rm = TRUE),
            mean_FRIP = mean(FRIP, na.rm = TRUE), .groups = "drop")

wilcox.test(FRIP ~ low_depth, data = frip_removed)

# FRIP vs depth across all samples, including the excluded ones
cor.test(frip_removed$reads_M, frip_removed$FRIP, method = "spearman")

ggplot(frip_removed, aes(reads_M, FRIP, colour = low_depth)) +
  geom_point(size = 1.5) +
  labs(x = "Paired reads (M)", y = "FRIP score (%)", colour = "Low depth")

  #isolate samples with FRIP < 60
low_frip <- frip_removed %>% filter(FRIP < 60)
low_frip

```